Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2025-02-15 demystify.tex
updates

\subsection{LoR from Well-Known People}  Having a strong letter from an internationally recognized researcher can \emph{greatly strengthen} your application. Such a letter can outweigh the lack of publications (or low GPA or unknown school). A well-known researcher has a reputation to maintain (so they won't recommend someone bad), and adcom member trust people that they know, or heard about and respect (compared to someone they don't know).

@@

  Many students get letters from supervisors from companies where they did internships or are working.  It is OK as long as it is a research-based personalized letter. Again, the emphasis here is \emph{research}, i.e., the letters should describe your research experiences and potential. Letters focusing on your coursework or non-research projects at a company won't carry much weight.

@@

\section{Work Experience} Work experience at \emph{well-known research laboratories}, such as Microsoft Research, can strengthen your
application.  The emphasis here is \emph{research} places, not software development or non-research work. For example, working at a FAANG (Facebook, Amazon, Apple, Netflix, Google) company as a software engineer does not count as research experience, although it can be helpful for MS applications discussed in \autoref{chap:ms}. Similarly, a LoR from your supervisor for non-research experience might not count much. So do not spend much time talking about development job in your SOP.

@@

If you do not get admitted to any schools or don't want to go to the ones that admit you, try again next time.  Graduate admission can involve randomness and noise. In the meantime, you can work on improving your profile, e.g., get more research experiences, publish more papers, improve your connections for stronger LoRs, etc. See rejection reasons (\autoref{sec:why-rejected}) and additional advice to improve your chance next time (\autoref{sec:improve-your-chance}).

@@

Many students lament that they get no interviews or are rejected and that the admission process seems random.  However, while it is true that the process is not perfect, it is not random.
Other than not having a strong profile (e.g., research potential, GPA, LoRs, SOP, poor interviews), here are some other reasons for rejections.

@@

There are other things you might not think are important but can make you stand out. For example, if you have a strong background in a non-CS field but can be integrated with CS, e.g., you have a degree or background in \emph{dance} or \emph{music} and want to integrate them with CS, do talk about it. Are you a female or a minority in CS? Do you participate in outreach activities that help increase diversity and inclusion in CS? Diversity is highly valued in CS programs in the US (\autoref{sec:urm}), and experiences in these areas can make you stand out and get noticed from reviewers.

>>>>

\subsection{LoR from Well-Known People}\label{sec:famous-lor} 
 Having a strong letter from an internationally recognized researcher can \emph{greatly strengthen} your application. Such a letter can outweigh the lack of publications (or low GPA or unknown school). A well-known researcher has a reputation to maintain (so they won't recommend someone bad), and adcom member trust people that they know, or heard about and respect (compared to someone they don't know).

@@

  Many students get letters from supervisors from companies where they did internships or are working.  It is OK as long as it is a research-based personalized letter. Again, the emphasis here is \emph{research}, i.e., the letters should describe your research experiences and potential. Letters focusing on non-research projects at a company won't carry much weight.

Finally, despite best intentions, the writers might not have the experience to write a strong LoR or lack the ability to evaluate your research ability.
This is unfortunate but common, and if you are in this situation, you should find someone else to write for you (see~\autoref{sec:lor-writers} and~\autoref{sec:famous-lor}).

@@

\section{Work Experience} Work experience at \emph{well-known research laboratories}, such as Microsoft Research, can strengthen your
application.  The emphasis here is \emph{research} places, not software development or non-research work. For example, working at a FAANG (Facebook, Amazon, Apple, Netflix, Google) company as a software engineer does not count as research experience, although it can be helpful for MS applications ()\autoref{chap:ms}). Similarly, a LoR from your supervisor for non-research experience might not count much (\autoref{sec:admin-letters}). So do not spend much time talking about development job in your SOP.

@@

If you do not get admitted to any schools or don't want to go to the ones that admit you, try again next year!  Grad. admission can involve randomness and noise. In the meantime, you can work on improving your profile, e.g., get more research experiences, publish more papers, improve your connections for stronger LoRs, etc. See rejection reasons (\autoref{sec:why-rejected}) and additional advice to improve your chance (\autoref{sec:improve-your-chance}).

@@

Many students lament that they get no interviews or are rejected and that the admission process seems random.  However, while it is true that the process is not perfect, it is not random.
It might be that your profile is not as strong as you \emph{think} it is.  Even if it is strong, you still can get rejected, e.g., not a good fit, overqualified, having redflags, etc.

\subsection{Your profile is not as strong as you think}  
I have seen multiple cases where students think they have a strong profile, but they do not. Here are some common examples.

Your \emph{LoRs} (\autoref{sec:lor-writers}) might not be as strong as you think. Your ref writers might not be very well-known, or active in research, or know how to write a good letter.

Your \emph{school} might not be as good as you think.  It might be T1 in your country (or in your own definition), but US faculty does not know about it, e.g., most US faculty would not know about schools in Vietnam.

Your \emph{research} might not be as strong as you think.  Doing research does not mean much (or anything) if there is no results, e.g,. publications. Publications might not have much value if they were published localled or through unknown places.  Even papers at top venues might not be as good as you think, e.g., many AI/ML/NLP faculty are not too interested in ``dataset'' papers at NeurIPS.

\subsection{Other Common Reasons for Rejection}
Even with strong profile (e.g., research potential, GPA, LoRs, SOP, interviews), you can still get rejected for various other reasons as follows.

@@

There are other things you might not think are important but can make you stand out. For example, if you have a strong background in a non-CS field but can be integrated with CS, e.g., you have a degree or background in \emph{dance} or \emph{music} and want to integrate them with CS, do talk about it. Are you a female or a minority in CS? Do you participate in outreach activities that help increase diversity and inclusion in CS? Diversity is valued in CS programs in the US (\autoref{sec:urm}), and experiences in these areas can make you stand out and get noticed from reviewers.
